### 2014-11-19 src/0_Abstract.tex
comenzando a juntar

\emph{Financial time series are known for their non-stationary behaviour. However,
sometimes they exhibit some stationary linear combinations. When this happens,
it is said that those time series are cointegrated. Vector error correction
model (VECM) is an econometric model which characterizes the joint dynamic
behaviour of a set of cointegrated variables in terms of forces pulling towards
equilibrium.}

@@

\emph{In this study, we propose an online VECM which optimizes how model parameters
are obtained by using only relevant past data and RR instead of OLS.  Our
proposal also takes advantage of the long-run relationship between the time
series in order to obtain improved execution times. Our proposed method is
tested using synthetic data and foreign exchange rates.
Results show that our proposal enables execution time to be reduced by at least
50\% while maintaining good accuracy levels.}

>>>>

\emph{Financial time series are known for their non-stationary behaviour and
this has motivated its study using different techniques. 
One relevant observation is that sometimes time series exhibit some stationary
linear combinations. When this happens, it is said that those time series are
cointegrated.  Cointegration has been then one of the main features studied.
Vector error correction model (VECM) is an econometric model which characterizes
the joint dynamic behaviour of a set of cointegrated variables in terms of
forces pulling towards equilibrium.}

@@

\emph{In this thesis, financial time series features are studied and different
algorithms were developed in order to optimize parameters an increase performance.
In particular, we propose an online VECM which optimizes how model parameters
are obtained considering only past data and different ways of solving the model
using machine learning techniques.  
The long-run relationship between the time series is used in order to make
optimizations and obtain improved execution times. However, due to the large
amount of data available and the need of quick response, algorithms presented
were optimized using high performance computing.
Our experiments were tested using synthetic data and foreign exchange rates.
Results show that cointegration and high performance computing allow to obtain
better models and execution times.}

### 2014-11-19 src/1_Introduction.tex
comenzando a juntar

\section{Introduction}
A time series is a sequence S of historical measurements y t of an observable
variable y at equal time intervals. Time series are studied for several purposes
such as the forecasting of the future based on knowledge of the past, the un-
derstanding of the phenomenon underlying the measures, or simply a succinct
description of the salient features of the series. In this chapter we shall
confine
ourselves to the problem of forecasting. Forecasting future values of an
observed
time series plays an important role in nearly all fields of science and
engineering,
such as economics, finance, business intelligence, meteorology and telecommu-
nication [43]. An important aspect of the forecasting task is represented by the
size of the horizon. If the one-step forecasting of a time series is already a
chal-
lenging task, performing multi-step forecasting is more difficult [53] because
of

additional complications, like accumulation of errors, reduced accuracy, and in-
creased uncertainty [58,49].
The forecasting domain has been influenced, for a long time, by linear sta-
tistical methods such as ARIMA models. However, in the late 1970s and early
1980s, it became increasingly clear that linear models are not adapted to many
real applications [25]. In the same period, several useful nonlinear time series
models were proposed such as the bilinear model [44], the threshold autoregres-
sive model [56,54,55] and the autoregressive conditional heteroscedastic (ARCH)
model [22] (see [25] and [26] for a review). However, the analytical study of
non-
linear time series analysis and forecasting is still in its infancy compared to
linear
time series [25].
In the last two decades, machine learning models have drawn attention and
have established themselves as serious contenders to classical statistical
models
in the forecasting community [1,43,61]. These models, also called black-box or
data-driven models [40], are examples of nonparametric nonlinear models which
use only historical data to learn the stochastic dependency between the past and
the future. For instance, Werbos found that Artificial Neural Networks (ANNs)
outperform the classical statistical methods such as linear regression and Box-
Jenkins approaches [59,60]. A similar study has been conducted by Lapedes and
Farber [33] who conclude that ANNs can be successfully used for modeling and
forecasting nonlinear time series. Later, other models appeared such as decision
trees, support vector machines and nearest neighbor regression [29,3]. Moreover,
the empirical accuracy of several machine learning models has been explored in a
number of forecasting competitions under different data conditions (e.g. the
NN3,
NN5, and the annual ESTSP competitions [19,20,34,35]) creating interesting
scientific debates in the area of data mining and forecasting [28,45,21].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The computational complexity of machine learning algorithms has become a limiting factor for problems that require processing large volumes of data and where the response time is crucial. An example of this is financial volatility time series forecast.

Currently, it is known that volatility is a process where many factors influence its behaviour. However, many models of volatility made forecasts based only on price returns. Examples are the ARFIMA, GARCH and stochastic volatility models. Machines learning methods have allowed a greater number of explanatory variables to be included, although in many cases, these increase calculation time.

Algorithms that process large amounts of data in a short time and are able to detect changes in the financial market are required to achieve fast and accurate responses. Online learning algorithms have been developed to solve large-scale problems since they process an instance at a time, updating the model at each step incrementally. This is opposed to the batch algorithms where a phase of training that uses a large collection of historical data is first required and the resulting model is later used to predict. 
This thesis proposal is to design and develop an online learning algorithm that meets or improves the performance of current volatility algorithms and ensures rapid responses to large-scale problems. Model comparisons will be made by the SPA (superior predictive ability) test proposed by Hansen and
Lunde~\cite{hansen+lunde2006}  that allows comparing of forecast models giving a relative measure of model performance with respect to one or more models.

\section{The problem}

In financial markets, volatility is one of the key elements to model the stochastic dynamic behaviour of financial assets. This is mainly because volatility gives a measure of uncertainty about future returns and it is often viewed as an indicator of the vulnerability of financial markets. It is also important as an input parameter in problems like derivative pricing, hedging and portfolio management. For example, in option pricing we need to known first the volatility of the underlying asset before pricing an option.

Besides, volatility provides important information for studying asset returns because the latter are largely uncorrelated and nonlinearly dependent. However, it has been found that volatility exhibits significant autocorrelation and predictable patterns~\cite{poon+granger2003} and  many models have been proposed to forecast its behaviour (see section ~\ref{sec:stylizedfacts}).

The volatility of the data is due to a large number of factors affecting the market, with some of them directly measurable such as historical prices, trends, supply and demand. However, others such as monetary policies and news are not directly measurable and are included in other models out of the scope of this thesis. Figure~\ref{fig:VolFactorNews} shows the VIX index, the daily US implied stock market volatility, commonly known as the "financial fear factor" which is clearly affected by historical events.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{img/vol-factor-thenews.jpg}
 % vol-factor-thenews.jpg: 710x506 pixel, 72dpi, 25.05x17.85 cm, bb=0 0 710 506
 \caption{The news as volatility factor}
 \label{fig:VolFactorNews}
\end{figure}

Volatility models have been extensively studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them include machine learning methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.  
The stochastic behaviour of volatility and the large number of factors affecting it result in models with a large number of parameters and slow response time due to the computationally expensive routines. These expensive calculations could be avoided using online learning algorithms, which allow response time reduction and are therefore suitable for high frequency trading applications.

In this thesis, a model will be proposed based on online learning algorithms to forecast volatility in high frequency financial time series.
The main problems to be deal with in this thesis are:

\begin{itemize}
   \item Although volatility is often predictable~\cite{poon+granger2003}, there are non-linear features~\cite{gatheral2006} to be considered in this proposal.
    \item The number of factors affecting volatility has been studied~\cite{srinivasan2009} and some other unknown factors affect volatility and they should also be included.
    \item The response time needs to be very fast and in order to ensure this, the proposed model will be presented as an online machine learning algorithm. 
\end{itemize}

\section{Scope of this Research}

Cointegration concept was introduced by~\cite{engle1987} and implies that one or
more linear combinations of non-stationary variables are stationary even though
individually they are not.  Moreover ~\cite{stock+watson1988} observed that
cointegration reflects the common stochastic trends providing a useful way to
understand cointegration relationships. These common stochastic trends can be
also interpreted as a long-run equilibrium relationships.

This idea was inmediatly adopted in finance since it could represent their
long-run relationship implied by economic theory~\cite{laietAl1991},
\cite{lence+falk2005}.  Economic theory suggest that economic time series are
mean-reverting process and therefore, it reflects the idea of that some set of
variables cannot wander too far from each other.

On the other hand, the efficient markets hypothesis, also known as the random
walk theory states that current stock prices fully reflect available
information related to its value and there is no way to earn excess
profits~\cite{fama1970}.  This means that if we have stock prices from a
jointly efficient market, they cannot be
cointegrated~\cite{granger1986},\cite{dwyer1992}. However, \cite{richards1995}
claims that cointegration is directly at odds with market efficiency, even
though, there is no evidende that cointegration among asset prices have
implications about market efficiency~\cite{lence+falk2005}.

Despite the fact that cointegration on closing daily rates of currency pairs has
not been found \cite{coleman1990}, \cite{copeland1991}, different time series
frequencies can have different behaviours~\cite{aldridge2009}. Pair trading is a
very common example of cointegration application~\cite{herlemont2003} 
but cointegration can also be extended to a larger set of
variables~\cite{mukherjee1995},\cite{engle2004}.

%So why do we care about cointegration? Someone else can probably give more
%econometric applications, but in quantitative finance, cointegration forms the
%basis of the pairs trading strategy: suppose we have two cointegrated stocks X
%and Y, with the particular (for concreteness) cointegrating relationship X - 2Y
%= Z, where Z is a stationary series of zero mean. For example, X could be
%McDonald's, Y could be Burger King, and the cointegration relationship would
%mean that X tends to be priced twice as high as Y, so that when X is more than
%twice the price of Y, we expect X to move down or Y to move up in the near
%future (and analogously, if X is less than twice the price of Y, we expect X to
%move up or Y to move down). This suggests the following trading strategy: if X -
%2Y > d, for some positive threshold d, then we should sell X and buy Y (since we
%expect X to decrease in price and Y to increase), and similarly, if X - 2Y < -d,
%then we should buy X and sell Y.

Vector error correction model (VECM) introduces this long-run relationship
among a set of cointegrated variables as an error correction term. VECM is a
special case of the vector autorregresive model (VAR) model. VAR model
expresses future values as a linear combination of variables past values.
However, VAR model cannot be used with non-stationary variables. VECM is a
linear model but in terms of variable differences. If cointegration exists,
variable differences are stationary and they introduce an error correction term
which adjusts coefficients to bring the variables back to equilibrium. In
finance, many economic time series turn to be stationary when they are
differentiated and cointegration restrictions often improves
forecasting~\cite{duy1998}. Therefore, VECM has been widely adopted.

Both VECM and VAR model parameters are obtained using ordinary least squares
(OLS) method. OLS has two main problems: is sensitive to errors on input data
and involves many calculations. The former problem is commonly solved using
Ridge Regression (RR) \cite{hoerl1970} which introduces a regularization
parameter that leads to an unbiased estimation with better generalization
capability. The second problem of computational complexity depends on the number
of past values and observations considered.  Recently, online learning
algorithms have been proposed to solve problems with large data sets because of
their simplicity and their ability to update the model when new data is
available.

@@

\section{Research Objectives}
The main motivation for this research is ...

The specific objectives of this research are,
\begin{itemize}
\item $\Large \mathcal{O}_1$: \emph{Especific Objective 1.}
\item $\Large \mathcal{O}_2$: \emph{Especific Objective 2.}
\item $\Large \mathcal{O}_3$: \emph{Especific Objective 3.}
\item $\Large \mathcal{O}_4$: \emph{Especific Objective 4.}
\item $\Large \mathcal{O}_5$: \emph{Especific Objective 5.}
\end{itemize}

>>>>

\section{Scope of this research}

The stochastic behaviour of financial time series, its incrementing
amount of data available and the need of performing accurate forecasting in
short periods of time has motivated researchers to create efficient and fast
algorithms. This study involves interdisciplinary knowledge such as: finance,
scientific computing, high performance computing, machine learning, physics
among others.

Forecasting financial time series have been modeled using classical statistical
models. More recently , machine learning models have been extensively used in
forecasting. However, getting model parameters is a computational challenge.
The computational complexity of machine learning algorithms has become a
limiting factor for problems that require processing large volumes of data and
where the response time is crucial.

Therefore, algorithms that process large amount of data in a short period of
time are required.  Online learning algorithms have been developed to solve
large-scale problems since they process an instance at a time, updating the
model at each step incrementally. This is opposed to the batch algorithms where
a phase of training that uses a large collection of historical data is first
required and the resulting model is later used to predict.

The specific scope of this study is to use financial time series features in
order to design a forecasting algorithm which ensures accuracy and low response
times. Cointegration is the main feature studied and it refers that one or more
linear combinations of these variables are stationary even though individually
they are not~\cite{engle87}.  Some models, such as the Vector Error Correction
(VECM), take advantage of this property and describe the joint behaviour of
several cointegrated variables.

@@

\section{Research Objectives}
The main motivation for this research is the develpment of efficient methods for
forecasting financial time series.

The specific objectives of this research are,
\begin{itemize}
\item $\Large \mathcal{O}_1$: \emph{A review of the literature on time series
analysis models including machine learning techniques.}
\item $\Large \mathcal{O}_2$: \emph{Development of a set of know features of the
studied time series and applied them to improve forecasting.}
\item $\Large \mathcal{O}_3$: \emph{Development of parallel and efficient
algorithms to ensure quick time responses.}
\item $\Large \mathcal{O}_4$: \emph{Deep mathematical analysis of the proposal
and financial concepts involved.}
\item $\Large \mathcal{O}_5$: \emph{Design and implement representative set of
experiments in order to show when and why proposal performs better.}
\end{itemize}
